### 2017-11-30 ident/docs/ident_manuscript.tex
changes to methods section

\begin{document}
\section{Introduction:}	
The dynamic response of any system can be modeled as a set of ordinary differential equations (ode) (Equation \ref{eq:output}a) in the state variables ($\mathbf{x}$). The changes in the state variables are expressed as a function of the state variables, the system parameters ($p$) and the inputs to the system ($u$).	
\begin{subequations}\label{eq:output}
	\begin{align}
	\dot{\mathbf{x}} = g(\mathbf{x}, p, u)\\
	\mathbf{y} = h(\mathbf{x},\mu,u)
	\end{align}
\end{subequations}
When not all states are observable, Equation (\ref{eq:output}a) is usually augmented with a system that defines the relationship between the observable output variables ($\mathbf{y}$) and the state variables ($\mathbf{x}$) as in Equation (\ref{eq:output}b). The parameters used to establish this relationship (Equation \ref{eq:output}b), $\mu$, may or may not include system parameters $p$ defined in Equation (\ref{eq:output}a).

Typically, a nonlinear programming formulation with the objective of minimizing the least square error between the measured ($y_{kl}^*$) and modeled ($y_{kl}$) outputs over the time course ($l = {1, ..., d}$) for which experimental data is collected (Equation \ref{eq:minnlp}) (Raue et al., and other parameter estimation/identifiability papers), is used to estimate parameters $p$ and $\mu$ in Equation \ref{eq:output}.
\begin{align}\label{eq:minnlp}
\chi^2(\theta) = \sum_{k=1}^{m}\sum_{l=1}^{d}\left(\frac{y_{kl}^*-y_{kl}}{\sigma_{kl}^*}\right)^2
\end{align}
The difference between the data and the model estimate for each output, and at each time point, is weighted by the variance in the experimental data $\sigma_{kl}^*$ for that corresponding variable and time point.
The minimization problem is solved subject to constraints on the state and output variables expressed in Equation \ref{eq:output}, as well as bounds on the estimated parameter values. The ability to determine a unique solution to the parameter estimation problem is however governed by the identifiability of the parameters in the model \parencite{McLean2012}.

@@

The dynamics of metabolic networks can also be represented by Equation (\ref{eq:output}) wherein the metabolite concentrations, $\mathbf{x}$, are state variables. The changes in the metabolite concentrations ($\mathbf{x}$) are expressed as a function of reaction rates dependent on enzyme kinetics, which are in turn expressed as nonlinear functions of the states ($\mathbf{x}$), kinetic parameters ($p$) and other input variables ($u$).

@@

@@

	\section{Methods:}
	\subsection{A method to establish posterior identifiability of metabolic network models:}
	This section details a method to establish the practical (posterior) identifiability of metabolic network models using the algebraic relationship between fluxes.
	Every flux, $v$, in a kinetic model of a metabolic network can be expressed as a nonlinear algebraic equation (Equation \ref{eq:flux_gen}). The fluxes are expressed as a function of the metabolite concentrations $x$ and the kinetic parameters $\theta$ in Equation (\ref{eq:flux_gen}).
	\begin{align}\label{eq:flux_gen}
	v = f(\mathbf{x},\mathbf{\theta})
	\end{align}
	Given the nonlinear nature of this model, the function $f$ in Equation (\ref{eq:flux_gen}) can expressed, without loss of generality as,
	\begin{align}\label{eq:nr_dr}
	v = \frac{N(\mathbf{x},\theta)}{D(\mathbf{x},\theta)}
	\end{align}
	where $N(\mathbf{x},\theta)$ is the numerator of $f$, and $D(\mathbf{x},\theta)$ is the denominator of $f$.

	If $\theta \in \mathbb{R}^p$, given a set of experimental measurements for the metabolite concentrations $\mathbf{x}$ and the reaction fluxes $\mathbf{v}$, theoretically, it is possible to choose $p$ sets of data from these measurements to solve for the $p$ parameters in $\theta$. However, if any these datasets do not satisfy the condition that $D(\mathbf{x},\theta) \neq 0$, then the number of experiments required to estimate the $p$ parameters in $\theta$ can be established to be greater than $p$. An example is shown below.

	This analysis can be performed for each flux in a metabolic network independent of all the other fluxes. This enables this method to be scalable to even genome-scale models. The following section demonstrates this methodology for one of the fluxes in the gluconeogenic model of Kotte et al., \parencite{Kotte2014}.

	\subsection{Identifiability analysis of parameters in a kinetic model of gluconeogenesis:}
	The proposed model for acetate consumption through gluconeogenesis and its corresponding kinetic model is used as a case study to illustrate the utility of identifiability analysis for the design of experiments for estimating parameters in kinetic models of metabolism. The kinetic model is described below.
	\begin{equation}\label{eq:ode1}
	\frac{d}{dt}pep=v_1-v_2-v_4
	\end{equation}
	\begin{equation}\label{eq:ode2}
	\frac{d}{dt}fdp=v_2-v_3
	\end{equation}
	\begin{equation}\label{eq:ode3}
	\frac{d}{dt}E=v_{e,max}\left(\frac{1}{1+\left(\frac{fdp}{K_{e}^{fdp}}\right)^{n_e}}\right) - d E
	\end{equation}
	The kinetic expressions for fluxes $v_1$ through $v_4$ are given below. The consumption of acetate through $v_1$ and conversion of \textit{pep} through $v_2$ are expressed in Equations (\ref{eq:flux1}) and (\ref{eq:flux2}) respectively using Michaelis-Menten kinetics. The acetate flux through $v_1$ is also governed by the quantity of available enzyme E. 
	\begin{equation}\label{eq:flux1}
	v_1 = k_{1}^{cat}E\frac{acetate}{acetate+K_{1}^{acetate}}
	\end{equation}	
	\begin{equation}\label{eq:flux2}
	v_2 = V_{2}^{max}\frac{pep}{pep+K_{2}^{pep}}
	\end{equation}
	\begin{equation}\label{eq:flux3}
	v_3 = V_{3}^{max}\frac{\tilde{fdp}\left(1+\tilde{fdp}\right)^3}{\left(1+\tilde{fdp}\right)^4+L_3\left(1+\frac{pep}{K_{3}^{pep}}\right)^{-4}}
	\end{equation}
	The allosterically regulated flux $v_3$ for the consumption of \textit{fdp} is expressed in Equation (\ref{eq:flux3}) using the Monod-Wyman-Changeux (MWC) model for allosterically regulated enzymes, where $\tilde{fdp}$ refers to the ratio of \textit{fdp} with respect to its allosteric binding constant $K_{3}^{fdp}$. The added flux $v_4$ for the export of \textit{pep} is expressed as a linear equation dependent on $pep$ in Equation (\ref{eq:flux4}).
	\begin{equation}\label{eq:flux4}
	v_4 = k_{4}^{cat}.pep
	\end{equation}

	We use flux $v_2$ to demonstrate the identifiability analysis method described in the previous section. Flux $v_2$ has two parameters, $V_2^{max}$ and $K_2^{pep}$ that need to be estimated from experimental data. Here, we assume that at least two different sets of experimental data for the concentrations and fluxes are available. Accordingly, we label these dataset as ${pep}^1$, ${v_2}^1$ and ${pep}^2$, ${v_2}^2$ respectively. Subsequently, these experimental datasets can be included in the model to form two simultaneous nonlinear algebraic equations in the parameters $V_2^{max}$ and $K_2^{pep}$ (Equation \ref{eq:v2_par}).

	\begin{subequations}\label{eq:v2_par}
		\begin{align}		
		V_2^{max} = \frac{v_2^1v_2^2(pep^1-pep^2)}{v_2^2pep^1-v_2^1pep^2}\\
		K_2^{pep} = \frac{pep^1(v_2^1pep^2-v_2^2pep^2)}{v_2^2pep^1-v_2^1pep^2}
		\end{align}
	\end{subequations}

@@

	\section{Results:}
	\paragraph{Outline:}
	\begin{itemize}
		\item parameter estimation is a well developed field typically using minimization of least square error to estimate model parameters from available experimental data
		\item if parameters are structurally identifiable, it does not guarantee practical identifiability from noisy experimental data
		\item identifiability dependent on whether given datasets (outputs) for estimation can sufficiently distinguish between different parameter values
	\end{itemize}

>>>>

\begin{document}
\section{Introduction:}	
The dynamic response of any system can be modeled as a set of ordinary differential equations (ode) (Equation \ref{eq:output}a) in the state variables ($\mathbf{x}$). The changes in the state variables are expressed as a function of the state variables, the system parameters ($\theta$) and the inputs to the system ($u$).	
\begin{subequations}\label{eq:output}
	\begin{align}
	\dot{\mathbf{x}} = g(\mathbf{x}, \theta, u)\\
	\mathbf{y} = h(\mathbf{x},\mu,u)
	\end{align}
\end{subequations}
When not all states are observable, Equation (\ref{eq:output}a) is usually augmented with a system that defines the relationship between the observable output variables ($\mathbf{y}$) and the state variables ($\mathbf{x}$) as in Equation (\ref{eq:output}b). The parameters used to establish this relationship (Equation \ref{eq:output}b), $\mu$, may or may not include system parameters $\theta$ defined in Equation (\ref{eq:output}a).

Typically, a nonlinear programming formulation with the objective of minimizing the least square error between the measured ($y_{kl}^*$) and modeled ($y_{kl}$) outputs over the time course ($l = {1, ..., d}$) for which experimental data is collected (Equation \ref{eq:minnlp}) (Raue et al., and other parameter estimation/identifiability papers), is used to estimate parameters $\theta$ and $\mu$ in Equation \ref{eq:output}.
\begin{align}\label{eq:minnlp}
\chi^2(\theta,\mu) = \sum_{k=1}^{m}\sum_{l=1}^{d}\left(\frac{y_{kl}^*-y_{kl}}{\sigma_{kl}^*}\right)^2
\end{align}
The difference between the data and the model estimate for each output, and at each time point, is weighted by the variance in the experimental data $\sigma_{kl}^*$ for that corresponding variable and time point.
The minimization problem is solved subject to constraints on the state and output variables expressed in Equation \ref{eq:output}, as well as bounds on the estimated parameter values. The ability to determine a unique solution to the parameter estimation problem is however governed by the identifiability of the parameters in the model \parencite{McLean2012}.

@@

The dynamics of metabolic networks can also be represented by Equation (\ref{eq:output}) wherein the metabolite concentrations, $\mathbf{x}$, are state variables. The changes in the metabolite concentrations ($\mathbf{x}$) are expressed as a function of reaction rates dependent on enzyme kinetics, which are in turn expressed as nonlinear functions of the states ($\mathbf{x}$), kinetic parameters ($\theta$) and other input variables ($u$).

@@

\section{Methods:}
\subsection{Kinetic model of gluconeogenesis in E. coli:}
The proposed model for acetate consumption through gluconeogenesis and its corresponding kinetic model is used as a case study to illustrate the utility of identifiability analysis for the design of experiments for estimating parameters in kinetic models of metabolism. The kinetic model is described below.
\begin{equation}\label{eq:ode1}
\frac{d}{dt}pep=v_1-v_2-v_4
\end{equation}
\begin{equation}\label{eq:ode2}
\frac{d}{dt}fdp=v_2-v_3
\end{equation}
\begin{equation}\label{eq:ode3}
\frac{d}{dt}E=v_{e,max}\left(\frac{1}{1+\left(\frac{fdp}{K_{e}^{fdp}}\right)^{n_e}}\right) - d E
\end{equation}
The kinetic expressions for fluxes $v_1$ through $v_4$ are given below. The consumption of acetate through $v_1$ and conversion of \textit{pep} through $v_2$ are expressed in Equations (\ref{eq:flux1}) and (\ref{eq:flux2}) respectively using Michaelis-Menten kinetics. The acetate flux through $v_1$ is also governed by the quantity of available enzyme E. 
\begin{equation}\label{eq:flux1}
v_1 = k_{1}^{cat}E\frac{acetate}{acetate+K_{1}^{acetate}}
\end{equation}	
\begin{equation}\label{eq:flux2}
v_2 = V_{2}^{max}\frac{pep}{pep+K_{2}^{pep}}
\end{equation}
\begin{equation}\label{eq:flux3}
v_3 = V_{3}^{max}\frac{\tilde{fdp}\left(1+\tilde{fdp}\right)^3}{\left(1+\tilde{fdp}\right)^4+L_3\left(1+\frac{pep}{K_{3}^{pep}}\right)^{-4}}
\end{equation}
The allosterically regulated flux $v_3$ for the consumption of \textit{fdp} is expressed in Equation (\ref{eq:flux3}) using the Monod-Wyman-Changeux (MWC) model for allosterically regulated enzymes, where $\tilde{fdp}$ refers to the ratio of \textit{fdp} with respect to its allosteric binding constant $K_{3}^{fdp}$. The added flux $v_4$ for the export of \textit{pep} is expressed as a linear equation dependent on $pep$ in Equation (\ref{eq:flux4}).
\begin{equation}\label{eq:flux4}
v_4 = k_{4}^{cat}.pep
\end{equation}

\subsection{A method for practical identifiability of kinetic models of metabolism:}
In this section, we show how practical identifiability of kinetic parameters in a dynamic model of metabolism can be established using the algebraic relationship between fluxes that form the network.
In a kinetic model, the value of every flux $v_i$ is expressed using one of the many available enzyme kinetic formulations. Without loss of generality, all of these kinetic formulations can be expressed as nonlinear algebraic equations (Equation \ref{eq:flux_gen}). The fluxes are expressed as a function of the metabolite concentrations $x$ and the kinetic parameters $\theta$. 
\begin{align}\label{eq:flux_gen}
v = f(\mathbf{x},\mathbf{\theta})=\frac{N(\mathbf{x},\theta)}{D(\mathbf{x},\theta)}
\end{align}
In Equation \ref{eq:flux_gen}, $N(\mathbf{x},\theta)$ is the numerator of $f$, and $D(\mathbf{x},\theta)$ is the denominator of $f$.

Let $\theta \in \mathbb{R}^p$ for each flux $v_i$ in the network. Also, let $E\in \mathbb{R}^n$ (and $j = {1, 2, ..., n}$) be the set of all possible experiments from which measurements of metabolite concentrations $\mathbf{x}_j$ and reaction fluxes $\mathbf{v}_j$ are available. For a given flux $v_i$, if all parameters $\theta \in \mathbb{R}^p$ are identifiable, then

if all parameters are identifiable, theoretically, the $P$ unknown parameters in $\theta$ can be estimated by solving $P\subset{1, 2, ..., n}$ nonlinear algebraic equations shown in Equation \ref{eq:nonlineq}.
\begin{align}\label{eq:nonlineq}
v_k = \frac{N(\mathbf{x}_k,\theta)}{D(\mathbf{x}_k,\theta)} && \forall k\in{1, 2, ..., k, ..., P}
\end{align}
Here, $K = {1, 2, ..., k, ..., P}$ is the index set of $P$ experiments for which metabolite concentration ($\mathbf{x}_K$) and flux ($\mathbf{v}_K$) measurements are available for each metabolite and flux within the metabolic network. Parameter $\theta_p\in\theta$ is said to be identifiable if

However, if for any index $l$, $\mathbf{x}_l$ and $\mathbf{v}_l$ do not satisfy the condition that $D(\mathbf{x}_l,\theta) \neq 0$, i.e., at least one flux $v_i$ is not identifiable based on experiment $l$, then the $l$th experiment cannot be used for parameter estimation and should be discarded. Consequently, an additional experiment $m \in {1, 2, ..., n}, m\neq l$ is needed to estimate all $p$ parameters in $\theta$. This analysis can be performed for each flux in a metabolic network independent of all the other fluxes. This enables this method to be scalable to even genome-scale models. The following section demonstrates this methodology for one of the fluxes in the gluconeogenic model described earlier.

\subsection{Experimental design for parameter estimation in kinetic models of metabolism:}
The identifiability of each parameter based on each experiment indexed as $j = {1, ..., n}$

\subsection{Identifiability of parameters in a kinetic model of gluconeogenesis:}

\subsection{Data for establishing parameter identifiability in kinetic model of gluconeogenesis:}
Steady state metabolomics and fluxomics data can be gathered under different physiological conditions by either perturbing the expression levels for different enzymes within a metabolic network, or by changing the substrate concentrations under which the cells grow. The aforementioned model of gluconeogenesis has three different fluxes ($v_1$, $v_2$ and $v_3$) whose enzyme expression parameters ($V_1^{max}$, $V_2^{max}$ and $V_3^{max}$) can be perturbed to simulate the repression and over expression of the corresponding enzymes. Furthermore, the acetate concentration on which the cell is grown can also be perturbed to measure cellular response to changes in the substrate concentration. We use the in silico metabolomics and fluxomics data generated from these perturbation experiments to demonstrate parameter identification with our methodology.

If any data generated from a perturbation experiment $i$ results in nonidentifiability, we eliminate experiment $i$ from the list of experiments that need to be performed for parameter estimation.

We use flux $v_2$ to demonstrate the identifiability analysis method described in the previous section. Flux $v_2$ has two parameters, $V_2^{max}$ and $K_2^{pep}$ that need to be estimated from experimental data. Here, we assume that at least two different sets of experimental data for the concentrations and fluxes are available. Accordingly, we label these dataset as ${pep}^1$, ${v_2}^1$ and ${pep}^2$, ${v_2}^2$ respectively. Subsequently, these experimental datasets can be included in the model to form two simultaneous nonlinear algebraic equations in the parameters $V_2^{max}$ and $K_2^{pep}$ (Equation \ref{eq:v2_par}).
\begin{subequations}\label{eq:v2_par}
	\begin{align}		
	V_2^{max} = \frac{v_2^1v_2^2(pep^1-pep^2)}{v_2^2pep^1-v_2^1pep^2}\\
	K_2^{pep} = \frac{pep^1(v_2^1pep^2-v_2^2pep^2)}{v_2^2pep^1-v_2^1pep^2}
	\end{align}
\end{subequations}

\section{Results:}
\subsection{Getting closed-form expressions for each flux in Kotte model:}
\subsection{Experiments that can establish identifiability for each flux in Kotte model:}	
\subsection{Combinations of experiments that will enable identification of all model parameters:}
\subsection{Expanding methodology to RBC model:}

\paragraph{Outline:}
\begin{itemize}
	\item parameter estimation is a well developed field typically using minimization of least square error to estimate model parameters from available experimental data
	\item if parameters are structurally identifiable, it does not guarantee practical identifiability from noisy experimental data
	\item identifiability dependent on whether given datasets (outputs) for estimation can sufficiently distinguish between different parameter values
\end{itemize}

@@

Given the nonlinear nature of this model, the function $f$ in Equation (\ref{eq:flux_gen}) can expressed, without loss of generality as,
\begin{align}\label{eq:nr_dr}
v = \frac{N(\mathbf{x},\theta)}{D(\mathbf{x},\theta)}
\end{align}

@@
